---
title: "How to use idaifieldR: Detailed Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use idaifieldR: Detailed Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  comment = "#>"
)
library(idaifieldR)
library(dplyr)
library(ggplot2)

# Load bundled test data so vignette builds without a live DB connection
idaifield_test_docs <- readRDS(system.file("testdata",
                                           "idaifield_test_docs.RDS",
                                           package = "idaifieldR"))
config <- readRDS(system.file("testdata",
                              "idaifield_test_config.RDS",
                              package = "idaifieldR"))
```

`idaifieldR` imports data from [iDAI.field / Field Desktop](https://github.com/dainst/idai-field) directly into R via the CouchDB API. This eliminates the need for CSV exports and allows scripts to be re-run and updated whenever the database changes.

## Installation

`idaifieldR` is not yet on CRAN. Install it from GitHub using `remotes`:

```{r eval = FALSE}
remotes::install_github("lsteinmann/idaifieldR")
library(idaifieldR)
```

## Connecting and Loading Data

To follow along with this tutorial, load the test dataset bundled with the
package into a new Field Desktop project called `rtest`. The import file is
located at
[`inst/testdata/rtest.jsonl`](https://github.com/lsteinmann/idaifieldR/blob/main/inst/testdata/rtest.jsonl)
on GitHub.


Create a connection object with `connect_idaifield()`:
```{r eval = FALSE}
conn <- connect_idaifield(pwd = "hallo", project = "rtest")
```

You can also ping the database to see if the connection succeeds:
```{r eval = FALSE}
idf_ping(conn)
```

If Field Desktop is running on the same computer, the `serverip` argument can
be left at its default (`"localhost"`). The `pwd` argument is the password set
in Field Desktop under *Tools > Settings > Your password*.

Load the complete project database into R:

```{r eval = FALSE}
idaifield_test_docs <- get_idaifield_docs(connection = conn)
```

The result is a named list of class `idaifield_docs` — one element per
resource, named by identifier, reflecting the original JSON structure. Loading
the whole database at once can be slow and memory-intensive for large projects;
see the [Queries] section for a more targeted approach.

## Building an Index

The index maps UUIDs to human-readable identifiers and is used by
`simplify_idaifield()` to make relations readable. Generate it with
`get_field_index()`:

```{r eval = FALSE}
index <- get_field_index(conn, verbose = TRUE, gather_trenches = TRUE)
```

```{r echo = FALSE}
index <- make_index(idaifield_test_docs, verbose = TRUE, gather_trenches = TRUE)
```

With `verbose = TRUE` the index also includes the `shortDescription`. With `gather_trenches = TRUE` the `Place` column is added, which contains the nearest *Place*-category resource (if any). Use `head(index)` to inspect it, or plot a quick overview of the dataset:

```{r indexPlot}
index |>
  ggplot(aes(y = Place, fill = category)) +
  geom_bar() +
  labs(title = "Resources by context and category", y = "Place", x = "Count")
```

## Unnesting (Optional)

`get_idaifield_docs()` returns the raw nested structure as it comes from the
API, containing the complete revision history for each resource. You can reduce 
it to the resource level with `maybe_unnest_docs()`:

```{r}
idaifield_test_resources <- maybe_unnest_docs(idaifield_test_docs)
```

This step is optional — all `idaifieldR` functions handle both
`idaifield_docs` and `idaifield_resources` lists. The unnested version takes
up less memory and is easier to browse in RStudio.

## Simplifying the Data

`simplify_idaifield()` transforms the nested list structure into something more
R-friendly:

- Relations are flattened into named vectors with a `relation.`-prefix
  (e.g. `relation.liesWithin`)
- UUIDs in relations are replaced with human-readable identifiers
- Geometry is kept as a GeoJSON string if `keep_geometry = TRUE`
- Single level lists are flattened to vectore, but more complex fields are preserved as-is.

You need to supply the `index` of the **complete** project database here, even
if you are working with a subset — it is used as a lookup table for UUID
replacement. You also need to supply the project `config`:

```{r eval = FALSE}
config <- get_configuration(conn)
```

```{r simplify, message = FALSE}
idaifield_test_simple <- simplify_idaifield(
  idaifield_test_resources,
  index        = index,
  config       = config,
  keep_geometry = TRUE
)
```

The result has class `idaifield_simple`. The connection and project name are
stored as attributes and carried through for later use. Check them with
`attributes(idaifield_test_simple)`.

## Selecting Resources

Use `idf_select_by()` to filter the simplified list by any field:

```{r}
pottery <- idf_select_by(idaifield_test_simple, by = "category", value = "Pottery")
```

## Queries

Alternatively, query the database directly instead of loading everything first.
This is useful for large projects where you only need a small subset:

```{r eval = FALSE}
pottery_docs <- idf_query(conn, field = "category", value = "Pottery")

# Always supply the index of the complete database when simplifying a subset
pottery_simple <- simplify_idaifield(pottery_docs, index = index, config = config, keep_geometry = TRUE)
```

See also `?idf_index_query` and `?idf_json_query` for more targeted queries.

## Converting to a Data Frame

Convert the simplified list to a matrix and then a data frame with
`idaifield_as_matrix()`:

```{r matrix}
pottery_df <- idaifield_as_matrix(pottery) |>
  as.data.frame()
```

From here, standard R and `ggplot2` workflows apply:

```{r potteryplot}
pottery_df |>
  ggplot(aes(x = relation.liesWithin, fill = period)) +
  geom_bar() +
  labs(title = "Pottery by context and period",
       x = "Context", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r potteryplot2}
pottery_df |>
  ggplot(aes(x = vesselForm, fill = specificType)) +
  geom_bar() +
  labs(title = "Pottery by vessel form and type",
       x = "Vessel form", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Geometry

Geometry is kept as a GeoJSON string when `keep_geometry = TRUE` in
`simplify_idaifield()`. To convert to an `sf` object for spatial analysis:

```{r eval = FALSE}
library(sf)
geom <- sf::st_read(pottery_simple$Befund_6_KER_4$geometry)
plot(geom)
```

Please note that you will need to set the CRS manually.

## Working with the Raw List

If the matrix format does not suit your analysis, work directly with the
`idaifield_resources` or `idaifield_simple` list. Each element is a named list
corresponding to one resource — extract fields directly, apply your own
`lapply()` transformations, or use `purrr` to reshape the data as needed.
