---
title: "TLDR: Essential Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TLDR: Essential Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 7,
  comment = "#>"
)
```

This vignette only highlights the most basic functions and shows a way to get to
your data quickly. 

## Connection: Example

```{r eval = FALSE}
conn <- connect_idaifield(pwd = "hallo", project = "rtest")
db <- get_idaifield_docs(connection = conn)
index <- get_uid_list(db)
```

## Selection and Simplify: Example

Using the following few lines and all default arguments while selecting from the whole database: 

```{r eval = FALSE}
library(dplyr)
pottery <- db %>%
  select_by(by = "category", value = "Pottery") %>%
  # Note: simplify_idaifield() needs the index of the complete project database 
  # as a lookup-table to replace all UUIDs with their identifiers
  simplify_idaifield(uidlist = index) %>%
  idaifield_as_matrix() %>%
  as.data.frame()
```
## Query and Simplify: Example

Or using the following few lines with direct queries to the database:

```{r eval = FALSE}
pottery <- idf_query(connection = conn,
                     field = "type", 
                     value = "Pottery") %>%
  simplify_idaifield(uidlist = index) %>%
  idaifield_as_matrix() %>%
  as.data.frame()
```

## Language Lookup: Example

Downloading translations and converting them into a lookup table:

```{r eval = FALSE}
core_lang_list <- download_language_list(project = "core", language = "de")
pergamon_lang_list <- download_language_list(project = "Pergamon", language = "de")
config_lang_list <- attr(db, "config")
config_lang_list <- config_lang_list$languages

lookup <- get_language_lookup(core_lang_list, language = "de")
lookup <- rbind(lookup, get_language_lookup(pergamon_lang_list, language = "de"))
lookup <- rbind(lookup, get_language_lookup(config_lang_list, language = "de"))

head(lookup)
```


```{r eval = FALSE}
colnames(pottery)

translated <- colnames(pottery)
names(translated) <- lookup$label[match(translated, lookup$var)]

translated
```



